---
title: "Metrics report"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: paper
    toc: true
    toc_float: true
    number_sections: true
params:
  functions_file: functions.R
  metrics_file: all-metrics.tsv
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(
    autodep        = TRUE,
    cache          = FALSE,
    cache.comments = FALSE,
    echo           = FALSE,
    error          = FALSE,
    message        = FALSE,
    warning        = FALSE,
    # dev            = "ragg_png",
    fig.align      = "center",
    # fig.width      = 960 / 72,
    fig.width      = 10,
    fig.asp        = 0.8,
    fig.retina     = 2,
    out.width      = "100%",
    crop           = TRUE,
    pngquant       = "--speed=1 --quality=0-50"
)
```

```{r libraries}
library(ggplot2)
library(readr)
library(dplyr)
```

```{r source}
source(params$functions_file)
```

```{r load_metrics}
metrics <- read_tsv(
    params$metrics_file,
    col_types = cols(
        .default = col_character(),
        Value    = col_double()
    )
) |>
    group_by(Dataset, Metric) |>
    mutate(ScaledValue = scales::rescale(Value)) |>
    ungroup()
```

# Summary

```{r metrics-summary}
tibble::tribble(
        ~ Item, ~ Number,
    "Datasets", length(unique(metrics$Dataset)),
     "Methods", length(unique(metrics$Method)),
     "Metrics", length(unique(metrics$Metric)),
      "Scores", nrow(metrics)
) |>
    knitr::kable()
```

# Integration metrics {.tabset}

```{r integration-dotplots, results = "hide"}
src_list <- lapply(unique(metrics$Dataset), function(.dataset) {
    src <- c(
        "## <<.dataset>> {.unnumbered .tabset .tabset-fade .tabset-pills}",
        "",
        "### Unscaled {.unnumbered}",
        "```{r integration-dotplots-<<.dataset>>-unscaled}",
        "plot_metrics_dotplot(",
        "metrics, '<<.dataset>>', metric_type = 'Integration', value = 'Value'",
        ")",
        "```",
        "",
        "### Scaled {.unnumbered}",
        "```{r integration-dotplots-<<.dataset>>-scaled}",
        "plot_metrics_dotplot(",
        "metrics, '<<.dataset>>', metric_type = 'Integration',",
        "value = 'ScaledValue'",
        ")",
        "```",
        ""
    )
    knitr::knit_expand(text = src, delim = c("<<", ">>"))
})

out <- knitr::knit_child(text = unlist(src_list), options = list(cache = FALSE))
```

`r out`

# Mapping metrics {.tabset}

```{r mapping-dotplots, results = "hide"}
src_list <- lapply(unique(metrics$Dataset), function(.dataset) {
    src <- c(
        "## <<.dataset>> {.unnumbered .tabset .tabset-fade .tabset-pills}",
        "",
        "### Unscaled {.unnumbered}",
        "```{r mapping-dotplots-<<.dataset>>-unscaled}",
        "plot_metrics_dotplot(",
        "metrics, '<<.dataset>>', metric_type = 'Mapping',",
        "value = 'Value'",
        ")",
        "```",
        "",
        "### Scaled {.unnumbered}",
        "```{r mapping-dotplots-<<.dataset>>-scaled}",
        "plot_metrics_dotplot(",
        "metrics, '<<.dataset>>', metric_type = 'Mapping',",
        "value = 'ScaledValue'",
        ")",
        "```",
        ""
    )
    knitr::knit_expand(text = src, delim = c("<<", ">>"))
})

out <- knitr::knit_child(text = unlist(src_list), options = list(cache = FALSE))
```

`r out`

# Classification metrics {.tabset}

```{r classification-dotplots, results = "hide"}
src_list <- lapply(unique(metrics$Dataset), function(.dataset) {
    src <- c(
        "## <<.dataset>> {.unnumbered .tabset .tabset-fade .tabset-pills}",
        "",
        "### Unscaled {.unnumbered}",
        "```{r classification-dotplots-<<.dataset>>-unscaled}",
        "plot_metrics_dotplot(",
        "metrics, '<<.dataset>>', metric_type = 'Classification',",
        "value = 'Value'",
        ")",
        "```",
        "",
        "### Scaled {.unnumbered}",
        "```{r classification-dotplots-<<.dataset>>-scaled}",
        "plot_metrics_dotplot(",
        "metrics, '<<.dataset>>', metric_type = 'Classification',",
        "value = 'ScaledValue'",
        ")",
        "```",
        ""
    )
    knitr::knit_expand(text = src, delim = c("<<", ">>"))
})

out <- knitr::knit_child(text = unlist(src_list), options = list(cache = FALSE))
```

`r out`

# Unseen population metrics {.tabset}

```{r unseen-dotplots, results = "hide"}
src_list <- lapply(unique(metrics$Dataset), function(.dataset) {
    src <- c(
        "## <<.dataset>> {.unnumbered .tabset .tabset-fade .tabset-pills}",
        "",
        "### Unscaled {.unnumbered}",
        "```{r unseen-dotplots-<<.dataset>>-unscaled}",
        "plot_metrics_dotplot(",
        "metrics, '<<.dataset>>', metric_type = 'Unseen',",
        "value = 'Value'",
        ")",
        "```",
        "",
        "### Scaled {.unnumbered}",
        "```{r unseen-dotplots-<<.dataset>>-scaled}",
        "plot_metrics_dotplot(",
        "metrics, '<<.dataset>>', metric_type = 'Unseen',",
        "value = 'ScaledValue'",
        ")",
        "```",
        ""
    )
    knitr::knit_expand(text = src, delim = c("<<", ">>"))
})

out <- knitr::knit_child(text = unlist(src_list), options = list(cache = FALSE))
```

`r out`
