---
title: "Number of features"
author: "Luke Zappia"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    fig.width = 10,
    fig.height = 8,
    fig.align = "center"
)
```

```{r libraries}
suppressPackageStartupMessages({
    library(tidyverse)
    library(patchwork)
})
```

```{r source}
source(here::here("analysis", "R", "plotting.R"))
source(here::here("analysis", "R", "summarisation.R"))
```

# Introduction {.unnumbered}

In this document we are going to perform analysis of the effect of the number of features for some common metric selection methods.

# Overview

First let's get an overview of the metric scores.

```{r load}
metrics <- read_tsv(
    here::here("data", "num-features.tsv"),
    col_types = cols(
        .default = col_character(),
        Value    = col_double()
    )
) |>
    mutate(
        Type = if_else(Metric == "GraphConnectivity", "IntegrationBio", Type)
    )

metrics_meta <- read_tsv(
    here::here("data", "metrics-metadata.tsv"),
    col_types = cols(
        Metric   = col_character(),
        Name     = col_character(),
        Included = col_logical()
    )
)
include <- metrics_meta$Metric[metrics_meta$Included]
exclude <- metrics_meta$Metric[!metrics_meta$Included]

metric_names <- metrics_meta$Name
names(metric_names) <- metrics_meta$Metric

datasets_meta <- read_tsv(
    here::here("data", "datasets-metadata.tsv"),
    col_types = cols(
        Dataset  = col_character(),
        Name     = col_character(),
    )
)

dataset_names <- datasets_meta$Name
names(dataset_names) <- datasets_meta$Dataset

type_names <- c(
    "IntegrationBatch" = "Integration (Batch)",
    "IntegrationBio" = "Integration (Bio)",
    "Mapping" = "Mapping",
    "Classification" = "Classification",
    "Unseen" = "Unseen populations"
)

metrics
```

## Counts

Check the number of values we have for each metric.
All metrics should have the same number (known missing values are checked below).

```{r counts}
plot_metric_counts(metrics)
```

## Missing values

We expect to have missing values for the `Reconstruction` metric which can't be calculated for the `Symphony` integration and the `CellCycle` metric which can't be calculated for simulated datasets.
Values should be complete for other metrics.

```{r missing}
plot_metric_missing(metrics)
```

# Summarise

Let's compute the summarised metrics using the baseline ranges.

```{r summarise}
baselines <- read_tsv(
    here::here("analysis", "output", "baseline-ranges.tsv"),
    col_types = cols(
        .default = col_character(),
        Lower    = col_double(),
        Upper    = col_double(),
        Range    = col_double()
    )
) |>
    mutate(
        Type = if_else(Metric == "GraphConnectivity", "IntegrationBio", Type)
    )

metrics_summary <- summarise_metrics(metrics, baselines)

metrics_summary
```

## Add number of features

For this analysis we are interested in the number of features so we need to split that into a separate column.
We also remove cases where the number of selected features is less than the number of features in the dataset and add the values for all features to each method.

```{r add-num-features}
metrics_summary <- metrics_summary |>
    mutate(
        SelFeatures = str_extract(Method, "N(\\d+)", group = 1),
        SelFeatures = as.numeric(SelFeatures),
        Method = str_remove(Method, "-N\\d+")
    ) |>
    left_join(
        select(datasets_meta, Dataset, Features),
        by = "Dataset"
    ) |>
    mutate(
        SelFeatures = if_else(Method == "all", Features, SelFeatures)
    ) |>
    filter(SelFeatures <= Features)

# Create rows for all features
metrics_summary_all <- metrics_summary |>
    filter(Method != "all") |>
    select(Dataset, Method, Integration) |>
    distinct() |>
    left_join(
        metrics_summary |>
            filter(Method == "all") |>
            select(-Method),
        by = c("Dataset", "Integration")
    )

metrics_summary <- metrics_summary |>
    filter(Method != "all") |>
    bind_rows(metrics_summary_all)

metrics_summary
```

# Visualisation

Let's visualise the effect of the number of features.

## Heatmap

The main heatmap shows the metric values for the example methods as the number of features increase, with the coloured rectangles highlighting the highest value in each row.
The bar chart above summarises for each metric type, how often the highest value is associated with each number of features.

```{r heatmap}
metrics_summary_plotting <- metrics_summary |>
    filter(Integration == "scVI-1", Method != "all") |>
    pivot_longer(
        cols = Classification:Overall,
        names_to = "Type",
        values_to = "Value"
    ) |>
    mutate(
        Dataset = factor(
            Dataset,
            levels = names(dataset_names),
            labels = dataset_names
        ),
        Dataset = str_replace(as.character(Dataset), " ", "\n"),
        Type = factor(
            Type,
            levels = c(names(type_names), "Overall"),
            labels = c(type_names, "Overall")
        ),
        Method = factor(
            Method,
            levels = c(
                "scanpy-cell_ranger-BatchFalse",
                "scanpy-pearson-BatchFalse",
                "scanpy-seurat-BatchFalse",
                "scanpy-seurat_v3-BatchFalse",
                "seurat-disp",
                "seurat-sct",
                "seurat-vst",
                "statistic-mean",
                "statistic-variance"
            ),
            labels = c(
                "scanpy-CellRanger",
                "scanpy-Pearson",
                "scanpy-Seurat",
                "scanpy-SeuratV3",
                "seurat-Dispersion",
                "seurat-scTransform",
                "seurat-VST",
                "statistic-Mean",
                "statistic-Variance"
            )
        ),
        SelFeatures = if_else(
            SelFeatures == Features,
            "All",
            as.character(SelFeatures)
        ),
        SelFeatures = factor(
            SelFeatures,
            levels = c(
                "100", "200", "500", "1000", "2000", "5000", "10000", "15000",
                "All"
            )
        )
    )

metrics_summary_top <- metrics_summary_plotting |>
    group_by(Dataset, Method, Integration, Type) |>
    slice_max(Value)

metrics_top_counts <- metrics_summary_top |>
    group_by(Type, SelFeatures) |>
    count(name = "Count")

types_palette <- c(RColorBrewer::brewer.pal(5, "Set1"), "#f781bf")

counts_barchart <- ggplot(
    metrics_top_counts,
    aes(x = SelFeatures, y = Count, fill = Type)
) +
    geom_col() +
    scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
    scale_fill_manual(values = types_palette) +
    facet_wrap(~ Type, nrow = 1) +
    labs(
        y = "Number of\ntop values"
    ) +
    theme_features(base_size = 10) +
    theme(
        legend.position = "none",
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        panel.grid.major.x = element_blank(),
    )

summary_heatmap <- ggplot(
    metrics_summary_plotting,
    aes(x = SelFeatures, y = Method, fill = Value)
) +
    geom_tile() +
    geom_tile(
        data = metrics_summary_top,
        aes(colour = Type),
        fill = NA, linewidth = 1
    ) +
    colorspace::scale_fill_continuous_diverging(
        palette = "Purple-Green"
    ) +
    scale_color_manual(values = types_palette, guide = "none") +
    facet_grid(Dataset ~ Type) +
    labs(
        x = "Number of features"
    ) +
    guides(
        custom = guide_custom(
            title = "Top\nvalue",
            grob = grid::rectGrob(
                height = unit(0.4, "cm"),
                width = unit(0.6, "cm")
            )
        )
    ) +
    theme_features(base_size = 10) +
    theme(
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        axis.title.y = element_blank(),
        panel.grid = element_blank(),
        strip.text.x = element_blank()
    )

num_features_plot <- wrap_plots(
    counts_barchart, summary_heatmap,
    ncol = 1,
    heights = c(1, 8),
    guides = "collect"
)

num_features_plot
```

## Line plot

To more clearly show the trend with the changing number of features we have standardised the values for each metric type by dataset and method.
This is visualised as line plots.

```{r line-plot}
metrics_summary_plotting <- metrics_summary_plotting |>
    group_by(Dataset, Method, Integration, Type) |>
    mutate(StandardValue = as.vector(scale(Value))) |>
    ungroup()

overall_means <- metrics_summary_plotting |>
    group_by(Type, SelFeatures) |>
    summarise(
        Value = mean(Value),
        StandardValue = mean(StandardValue),
        .groups = "drop"
    )

overall_plot <- ggplot(
    metrics_summary_plotting,
    aes(x = SelFeatures, y = StandardValue, colour = Type, fill = Type)
) +
    geom_hline(yintercept = 0, colour = "red") +
    ggforce::geom_sina(alpha = 0.2) +
    geom_line(
        data = filter(overall_means, SelFeatures != "All"),
        aes(group = Type),
        linewidth = 1
    ) +
    geom_point(
        data = overall_means,
        shape = "diamond filled", size = 3, stroke = 1, fill = "white"
    ) +
    scale_color_manual(values = types_palette, guide = "none") +
    scale_fill_manual(values = types_palette, guide = "none") +
    facet_grid(. ~ Type) +
    labs(
        y = "Standardised value"
    ) +
    theme_features() +
    theme(
        legend.position = "none",
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        panel.grid = element_blank(),
    )

dataset_means <- metrics_summary_plotting |>
    group_by(Dataset, Type, SelFeatures) |>
    summarise(
        Value = mean(Value),
        SD = sd(StandardValue),
        StandardValue = mean(StandardValue),
        .groups = "drop"
    )

datasets_plot <- ggplot(
    metrics_summary_plotting,
    aes(x = SelFeatures, y = StandardValue, colour = Type, fill = Type)
) +
    geom_hline(yintercept = 0, colour = "red") +
    ggforce::geom_sina(alpha = 0.2) +
    geom_line(
        data = filter(dataset_means, SelFeatures != "All"),
        aes(group = Type),
        linewidth = 0.5
    ) +
    geom_point(
        data = dataset_means,
        shape = "diamond filled", size = 2, stroke = 0.5, fill = "white"
    ) +
    scale_color_manual(values = types_palette, guide = "none") +
    scale_fill_manual(values = types_palette, guide = "none") +
    facet_grid(Dataset ~ Type) +
    labs(
        y = "Standardised value"
    ) +
    theme_features() +
    theme(
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        panel.grid = element_blank(),
        strip.text.x = element_blank()
    )

wrap_plots(
    overall_plot, datasets_plot,
    ncol = 1,
    heights = c(2, 8),
    guides = "collect"
)
```

# Summary

This figure summarised the effect of the number of selected features on metric values.

```{r summary}
overall_lineplot <- ggplot(
    metrics_summary_plotting,
    aes(x = SelFeatures, y = StandardValue, colour = Type, fill = Type)
) +
    geom_hline(yintercept = 0, colour = "red") +
    ggforce::geom_sina(size = 0.5, alpha = 0.2) +
    geom_line(
        data = filter(overall_means, SelFeatures != "All"),
        aes(group = Type),
        linewidth = 1
    ) +
    geom_point(
        data = overall_means,
        shape = "diamond filled", size = 1.6, stroke = 0.6, fill = "white"
    ) +
    scale_color_manual(values = types_palette, guide = "none") +
    scale_fill_manual(values = types_palette, guide = "none") +
    facet_grid(. ~ Type) +
    labs(
        y = "Standardised value"
    ) +
    theme_features_pub() +
    theme(
        legend.position = "none",
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        panel.grid = element_blank(),
    )

method_means <- metrics_summary_plotting |>
    group_by(Method, Type, SelFeatures) |>
    summarise(
        Value = mean(Value),
        SD = sd(StandardValue),
        StandardValue = mean(StandardValue),
        .groups = "drop"
    )
mean_limits <- c(
    min(method_means$StandardValue, dataset_means$StandardValue),
    max(method_means$StandardValue, dataset_means$StandardValue)
)

datasets_heatmap <- ggplot(
    mutate(dataset_means, Dataset = str_replace(Dataset, "\n", " ")),
    aes(x = SelFeatures, y = Dataset, colour = StandardValue, size = SD)
) +
    geom_point(shape = "square") +
    colorspace::scale_colour_continuous_diverging(
        palette = "Purple-Green",
        limits = mean_limits
    ) +
    scale_size_continuous(
        trans = "reverse",
        limits = c(max(dataset_means$SD), 0),
        range = c(0.1, 2.5)
    ) +
    facet_grid(. ~ Type) +
    labs(
        title = "Datasets",
        x = "Number of selected features",
        colour = "Mean standardised value",
        size = "Standard deviation of\nstandardised values"
    ) +
    guides(
        colour = guide_colourbar(order = 1),
        size = guide_legend(
            theme = theme(legend.text.position = "bottom"),
            order = 2
        )
    ) +
    theme_features_pub() +
    theme(
        legend.position = "bottom",
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.title.y = element_blank(),
        panel.grid = element_blank(),
        strip.background = element_blank(),
        strip.text = element_blank()
    )

methods_heatmap <- ggplot(
    method_means,
    aes(x = SelFeatures, y = Method, colour = StandardValue, size = SD)
) +
    geom_point(shape = "square") +
    colorspace::scale_colour_continuous_diverging(
        palette = "Purple-Green",
        limits = mean_limits
    ) +
    scale_size_continuous(
        trans = "reverse",
        limits = c(max(dataset_means$SD), 0),
        range = c(0.1, 2.5)
    ) +
    facet_grid(. ~ Type) +
    labs(
        title = "Methods",
        x = "Number of selected features",
        colour = "Mean standardised value",
        size = "Standard deviation of\nstandardised values"
    ) +
    guides(
        colour = guide_colourbar(order = 1),
        size = guide_legend(
            theme = theme(legend.text.position = "bottom"),
            order = 2
        )
    ) + 
    theme_features_pub() +
    theme(
        legend.position = "bottom",
        axis.title.y = element_blank(),
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        panel.grid = element_blank(),
        strip.background = element_blank(),
        strip.text.x = element_blank()
    )

summary_plot <- wrap_plots(
    overall_lineplot, datasets_heatmap, methods_heatmap,
    ncol = 1,
    heights = c(1, 0.7, 0.7),
    guides = "collect"
) &
    theme(
        legend.position = "bottom",
        legend.title.position = "top"
    )

summary_plot
```

# Output

Save output files

```{r output}
save_figure_files(
    summary_plot,
    here::here("analysis", "output", "num-features"),
    height = 5.5
)
```

# Session info {.unnumbered}

```{r session-info}
sessioninfo::session_info()
```

