---
title: "HLCA lineages"
author: "Luke Zappia"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    fig.width = 10,
    fig.height = 8,
    fig.align = "center"
)
```

```{r libraries}
suppressPackageStartupMessages({
    library(tidyverse)
    library(patchwork)
})
```

```{r source}
source(here::here("analysis", "R", "plotting.R"))
source(here::here("analysis", "R", "summarisation.R"))
```

# Introduction {.unnumbered}

In this document we are going to perform analysis of three versions of the HLCA dataset.

# Overview

First let's get an overview of the metric scores.

```{r load}
metrics <- read_tsv(
    here::here("data", "benchmark.tsv"),
    col_types = cols(
        .default = col_character(),
        Value    = col_double()
    )
) |>
    mutate(
        Type = if_else(Metric == "GraphConnectivity", "IntegrationBio", Type)
    )

metrics_meta <- read_tsv(
    here::here("data", "metrics-metadata.tsv"),
    col_types = cols(
        Metric   = col_character(),
        Name     = col_character(),
        Included = col_logical()
    )
)
include <- metrics_meta$Metric[metrics_meta$Included]
exclude <- metrics_meta$Metric[!metrics_meta$Included]

metric_names <- metrics_meta$Name
names(metric_names) <- metrics_meta$Metric

datasets_meta <- read_tsv(
    here::here("data", "datasets-metadata.tsv"),
    col_types = cols(
        Dataset  = col_character(),
        Name     = col_character(),
    )
)

dataset_names <- datasets_meta$Name
names(dataset_names) <- datasets_meta$Dataset

type_names <- c(
    "IntegrationBatch" = "Integration (Batch)",
    "IntegrationBio" = "Integration (Bio)",
    "Mapping" = "Mapping",
    "Classification" = "Classification",
    "Unseen" = "Unseen populations"
)

metrics
```

## Counts

Check the number of values we have for each metric.
All metrics should have the same number (known missing values are checked below).

```{r counts}
plot_metric_counts(metrics)
```

## Missing values

We expect to have missing values for the `Reconstruction` metric which can't be calculated for the `Symphony` integration and the `CellCycle` metric which can't be calculated for simulated datasets.
Values should be complete for other metrics.

```{r missing}
plot_metric_missing(metrics)
```

# Summarise

Let's compute the summarised metrics using the baseline ranges.

```{r summarise}
baselines <- read_tsv(
    here::here("analysis", "output", "baseline-ranges.tsv"),
    col_types = cols(
        .default = col_character(),
        Lower    = col_double(),
        Upper    = col_double(),
        Range    = col_double()
    )
) |>
    mutate(
        Type = if_else(Metric == "GraphConnectivity", "IntegrationBio", Type)
    )

metrics_summary <- metrics |>
    # Calculate average scores for random methods
    mutate(
        Method = str_remove(Method, "-S[0-9]+$"),
    ) |>
    group_by(Dataset, Method, Integration, Type, Metric) |>
    summarise(Value = mean(Value), .groups = "drop") |>
    # Scale and summarise
    summarise_metrics(baselines)

metrics_summary
```

## Filter

For this analysis we are only interested in the HLCA datasets so we will filter to keep only those datasets and only the scVI integration.

```{r filter}
metrics_summary_all <- metrics_summary

metrics_summary <- metrics_summary |>
    filter(
        Dataset %in% c("HLCA", "HLCAImmune", "HLCAEpithelial"),
        Integration == "scVI-1"
    )
```

# Ranks

Let's compute the method ranks for each dataset and compare them.

```{r ranks}
metrics_ranks <- metrics_summary |>
    pivot_longer(
        Classification:Overall,
        names_to = "Type",
        values_to = "Value"
    ) |>
    group_by(Dataset, Type) |>
    mutate(
        Rank = rank(Value)
    )

ggplot(metrics_ranks, aes(x = Dataset, y = Method, fill = Rank)) +
    geom_tile() +
    facet_wrap(~ Type, nrow = 1) +
    scale_fill_viridis_c() +
    theme_features() +
    theme(
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
    )
```

# Overlap

Here we compare the overlap in selected features between the dataset subsets.

```{r overlaps}
feature_files <- fs::dir_ls(
    here::here("data", "selected-features"),
    glob = "*HLCA*.tsv"
)

selected_list <- map(feature_files, function(.file) {
    read_tsv(
        .file,
        col_types = cols(
            .default = col_logical(),
            Feature = col_character()
        )
    ) |>
        column_to_rownames("Feature") |>
        as.matrix()
})
names(selected_list) <- fs::path_ext_remove(fs::path_file(names(selected_list)))

datasets_grid <- expand_grid(
    dataset1 = factor(names(selected_list)),
    dataset2 = factor(names(selected_list))
) |>
    filter(as.numeric(dataset1) < as.numeric(dataset2))

overlaps <- map_dfr(colnames(selected_list[[1]]), function(.method) {
    pmap_dfr(datasets_grid, function(dataset1, dataset2) {
        
        selected1 <- selected_list[[dataset1]][, .method]
        selected2 <- selected_list[[dataset2]][, .method]
        
        selected_names1 <- rownames(selected_list[[dataset1]])[selected1]
        selected_names2 <- rownames(selected_list[[dataset2]])[selected2]
        
        both <- length(intersect(selected_names1, selected_names2))
        either <- length(union(selected_names1, selected_names2))
        
        tibble(
            Method = .method,
            Dataset1 = dataset1,
            Dataset2 = dataset2,
            Both = both,
            Either = either,
            Jaccard = both / either,
        )
    })
})

overlaps |>
    filter(str_detect(Method, "random", negate = TRUE)) |>
    mutate(Combination = paste(Dataset1, Dataset2, sep = " - ")) |>
    ggplot(aes(x = Combination, y = Method, fill = Jaccard)) +
        geom_tile() +
        scale_fill_viridis_c() +
        theme_features() +
        theme(
            axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
        )
```

# Markers overlap

Now we will look at the overlap between selected features and marker gene sets.

```{r markers-overlap}
marker_overlaps <- read_tsv(
    here::here("data", "hlca-marker-overlaps.tsv"),
    col_types = cols(
        method = col_character(),
        dataset = col_character(),
        cell_type = col_character(),
        lineage = col_character(),
        score = col_double()
    )
) |>
    rename(
        Method = method,
        Dataset = dataset,
        Label = cell_type,
        Lineage = lineage,
        Prop = score
    )

marker_overlap_means <- marker_overlaps |>
    group_by(Method, Dataset, Lineage) |>
    summarise(
        MeanProp = mean(Prop),
        SDProp = sd(Prop),
        .groups = "drop"
    )

ggplot(
    filter(marker_overlap_means, str_detect(Method, "random", negate = TRUE)),
    aes(x = Dataset, y = Method, colour = MeanProp, size = SDProp)
) +
    geom_point(shape = "square") +
    facet_wrap(~ Lineage, nrow = 1) +
    scale_colour_viridis_c() +
    scale_size_continuous(trans = "reverse") +
    theme_features() +
    theme(
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
    )
```

# Label Milo

To directly compare the different datasets we consider Milo scores calculated for each unseen label.

## Scores

Milo scores for each unseen label.

```{r label-Milo}
label_milo <- read_tsv(
    here::here("data", "hlca-label-MILO.tsv"),
    col_types = cols(
        Method = col_character(),
        Dataset = col_character(),
        Integration = col_character(),
        Type = col_character(),
        Metric = col_character(),
        Label = col_character(),
        Value = col_double()
    )
) |>
    filter(Integration == "scVI-1") |>
    # Calculate average scores for random methods
    mutate(
        Method = str_remove(Method, "-S[0-9]+$"),
    ) |>
    group_by(Dataset, Method, Integration, Type, Metric, Label) |>
    summarise(Value = mean(Value), .groups = "drop")

ggplot(label_milo, aes(x = Label, y = Method, fill = Value)) +
    geom_tile() +
    facet_wrap(~ Dataset, nrow = 1) +
    scale_fill_viridis_c() +
    theme_features() +
    theme(
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
    )
```

## Differences

Differences in Milo scores compared to full dataset.

```{r label-Milo-diff}
label_milo_diff <- label_milo |>
    pivot_wider(names_from = Dataset, values_from = Value) |>
    mutate(
        EpithelialDiff = HLCAEpithelial - HLCA,
        ImmuneDiff = HLCAImmune - HLCA
    ) |>
    select(
        -HLCA,
        -HLCAEpithelial,
        -HLCAImmune
    ) |>
    pivot_longer(
        EpithelialDiff:ImmuneDiff,
        names_to = "Dataset",
        values_to = "Difference"
    ) |>
    filter(!is.na(Difference))

ggplot(label_milo_diff, aes(x = Label, y = Method, fill = Difference)) +
    geom_tile() +
    colorspace::scale_fill_continuous_diverging(
        palette = "Purple-Green",
        limits = c(-1, 1)
    ) +
    facet_wrap(~ Dataset, nrow = 1) +
    theme_features() +
    theme(
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
    )
```

# Figures

Final figures for the lineages analysis.

```{r figure}
type_names <- c(
    "IntegrationBatch" = "Integration (Batch)",
    "IntegrationBio" = "Integration (Bio)",
    "Mapping" = "Mapping",
    "Classification" = "Classification",
    "Unseen" = "Unseen populations"
)

types_palette <- c("#f781bf", RColorBrewer::brewer.pal(5, "Set1"))

methods_meta <- read_tsv(
    here::here("data", "methods-metadata.tsv"),
    col_types = cols(
        Method     = col_character(),
        Name       = col_character(),
        IsBaseline = col_logical()
    )
)
method_names <- methods_meta$Name
names(method_names) <- methods_meta$Method

methods_order <- metrics_summary_all |>
    filter(Integration == "scVI-1") |>
    pivot_longer(
        cols = Classification:Overall,
        names_to = "Type",
        values_to = "Value"
    ) |>
    group_by(Method, Type) |>
    summarise(Value = mean(Value), .groups = "drop") |>
    filter(Type == "Overall") |>
    arrange(Value) |>
    pull(Method)

metrics_ranks_plotting <- metrics_ranks |>
    mutate(
        Method = factor(
            Method,
            levels = methods_order,
            labels = method_names[methods_order]
        ),
        Type = factor(
            Type,
            levels = c("Overall", names(type_names)),
            labels = c("Overall", str_replace(type_names, " ", "\n"))
        ),
        Dataset = factor(
            Dataset,
            levels = c("HLCA", "HLCAImmune", "HLCAEpithelial"),
            labels = c("Full", "Immune", "Epithelial")
        )
    )

ranks_plot <- ggplot(
    metrics_ranks_plotting,
    aes(
        x = Dataset, y = Method, fill = Type,
        alpha = length(unique(Method)) -  Rank)
) +
    geom_tile() +
    facet_wrap(~ Type, nrow = 1) +
    scale_x_discrete(expand = c(0, 0)) +
    scale_y_discrete(expand = c(0, 0), drop = FALSE) +
    scale_fill_manual(values = types_palette, guide = "none") +
    scale_alpha_continuous(
        limits = c(length(unique(metrics_ranks_plotting$Method)), 1),
        breaks = c(0, 5, 10, 15, 20, 25, 30),
        range = c(0.1, 1.0),
        trans = "reverse"
    ) +
    labs(
        title = "Method ranks for HLCA subsets",
        alpha = "Rank"
    ) +
    theme_features_pub() +
    theme(
        legend.position = "bottom",
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        axis.title = element_blank(),
        panel.grid.major = element_blank(),
    )

overlaps_plotting <- overlaps |>
    filter(str_detect(Method, "random", negate = TRUE)) |>
    mutate(
        Method = Method |>
            str_replace_all("_", "-") |>
            str_replace("seurat-v3", "seurat_v3") |>
            str_replace("cell-ranger", "cell_ranger") |>
            factor(
                levels = methods_order,
                labels = method_names[methods_order]
            ),
        Dataset1 = factor(
            Dataset1,
            levels = c("HLCA", "HLCAImmune", "HLCAEpithelial"),
            labels = c("Full", "Immune", "Epithelial")
        ),
        Dataset2 = factor(
            Dataset2,
            levels = c("HLCA", "HLCAImmune", "HLCAEpithelial"),
            labels = c("Full", "Immune", "Epithelial")
        )
    ) |>
    mutate(
        Combination = paste(Dataset1, Dataset2, sep = " vs "),
        Combination = factor(
            Combination,
            levels = c(
                "Full vs Immune",
                "Full vs Epithelial",
                "Epithelial vs Immune"
            )
        )
    )
    
overlaps_plot <- ggplot(
    overlaps_plotting, aes(x = Combination, y = Method, fill = Jaccard)
) +
    geom_tile() +
    scale_fill_viridis_c(option = "plasma", limits = c(0, 1)) +
    scale_x_discrete(expand = c(0, 0)) +
    scale_y_discrete(expand = c(0, 0), drop = FALSE) +
    labs(
        title = "Selected\nfeatures",
        fill = "Jaccard index"
    ) +
    theme_features_pub() +
    theme(
        legend.position = "bottom",
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        axis.text.y = element_blank(),
        axis.title = element_blank(),
        panel.grid = element_blank()
    )

marker_overlap_means_plotting <- marker_overlap_means |>
    filter(
        Method != "all-features",
        str_detect(Method, "random", negate = TRUE)
    ) |>
    mutate(
        Method = Method |>
            str_replace_all("_", "-") |>
            str_replace("seurat-v3", "seurat_v3") |>
            str_replace("cell-ranger", "cell_ranger") |>
            factor(
                levels = methods_order,
                labels = method_names[methods_order]
            ),
        Dataset = factor(
            Dataset,
            levels = c("HLCA", "HLCAImmune", "HLCAEpithelial"),
            labels = c("Full", "Immune", "Epithelial")
        )
    )

markers_overlap_plot <- ggplot(
    marker_overlap_means_plotting,
    aes(x = Dataset, y = Method, colour = MeanProp, size = SDProp)
) +
    geom_point(shape = "square") +
    facet_wrap(~ Lineage, nrow = 1) +
    scale_y_discrete(drop = FALSE) +
    scale_colour_viridis_c() +
        scale_size_continuous(
        trans = "reverse",
        limits = c(max(marker_overlap_means_plotting$SDProp), 0),
        range = c(0.1, 2.5)
    ) +
    labs(
        title = "Overlap with marker gene sets",
        colour = "Mean proportion of markers",
        size = "SD proportion of markers"
    ) +
    theme_features_pub() +
    theme(
        legend.position = "bottom",
        axis.title = element_blank(),
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        axis.text.y = element_blank(),
        panel.grid.major = element_blank()
    )

top_panel <- wrap_plots(
    ranks_plot,
    overlaps_plot,
    markers_overlap_plot,
    nrow = 1,
    widths = c(1, 0.1, 0.6),
    guides = "collect"
) &
    theme(
        legend.position = "bottom",
        legend.title.position = "top"
    )

labels_clust <- label_milo |>
    filter(Dataset == "HLCA") |>
    select(Method, Label, Value) |>
    pivot_wider(names_from = Label, values_from = Value) |>
    column_to_rownames("Method") |>
    as.matrix() |>
    t() |>
    dist() |>
    hclust()

# Arrange labels by lineage then clustering order
labels_order <- label_milo |>
    mutate(
        Lineage = case_when(
            Label %in%
                (filter(label_milo, Dataset == "HLCAImmune") |>
                pull(Label) |> unique()) ~
                "Immune",
            Label %in%
                (filter(label_milo, Dataset == "HLCAEpithelial") |>
                pull(Label) |> unique()) ~
                "Epithelial",
            TRUE ~ "Other"
        ),
        Lineage = factor(
            Lineage,
            levels = c("Immune", "Epithelial", "Other")
        ),
        Label = factor(
            Label,
            levels = labels_clust$labels[labels_clust$order]
        )
    ) |>
    filter(Dataset == "HLCA") |>
    arrange(Lineage, Label) |>
    pull(Label) |>
    as.character() |>
    unique()

label_milo_plotting <- label_milo |>
    mutate(
        Method = factor(
            Method,
            levels = methods_order,
            labels = method_names[methods_order]
        ),
        Dataset = factor(
            Dataset,
            levels = c("HLCA", "HLCAImmune", "HLCAEpithelial"),
            labels = c("Full", "Immune", "Epithelial")
        ),
        Label = factor(
            Label,
            levels = labels_order
        )
    )

milo_plot <- ggplot(
    label_milo_plotting,
    aes(x = Label, y = Method, fill = Value)
) +
    geom_tile() +
    facet_grid(. ~ Dataset, scales = "free_x", space = "free_x") +
    scale_fill_viridis_c(option = "magma") +
    scale_x_discrete(expand = c(0, 0)) +
    scale_y_discrete(expand = c(0, 0), drop = FALSE) +
    labs(
        title = "Unssen cell label MILO scores",
        fill = "Milo score"
    ) +
    theme_features_pub() +
    theme(
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        axis.title = element_blank(),
        panel.grid = element_blank()
    )

label_milo_diff_plotting <- label_milo_diff |>
    mutate(
        Method = factor(
            Method,
            levels = methods_order,
            labels = method_names[methods_order]
        ),
        Dataset = factor(
            Dataset,
            levels = c("ImmuneDiff", "EpithelialDiff"),
            labels = c("Immune", "Epithelial")
        ),
        Label = factor(
            Label,
            levels = labels_order
        )
    )

milo_diff_plot <- ggplot(
    label_milo_diff_plotting,
    aes(x = Label, y = Method, fill = Difference)
) +
    geom_tile() +
    colorspace::scale_fill_continuous_diverging(
        palette = "Purple-Green",
        limits = c(-1, 1)
    ) +
    scale_x_discrete(expand = c(0, 0)) +
    scale_y_discrete(expand = c(0, 0), drop = FALSE) +
    facet_grid(. ~ Dataset, scales = "free_x", space = "free_x") +
    labs(
        title = "Difference in scores"
    ) +
    theme_features_pub() +
    theme(
        axis.title = element_blank(),
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        axis.text.y = element_blank(),
        panel.grid = element_blank()
    )

milo_panel <- wrap_plots(
    milo_plot,
    milo_diff_plot,
    nrow = 1,
    widths = c(1, 0.4),
    guides = "collect"
) &
    theme(
        legend.position = "bottom",
        legend.title.position = "top"
    )

lineages_figure <- wrap_plots(
    top_panel,
    milo_panel,
    heights = c(1.0, 0.9),
    nrow = 2
)

lineages_figure
```

# Supplementary

Extended figures for supplementary materials.

## Marker overlaps

```{r supplementary-overlaps, fig.height = 10, out.width = "100%"}
marker_overlaps_plotting <- marker_overlaps |>
    filter(
        Method != "all-features",
        str_detect(Method, "random", negate = TRUE)
    ) |>
    mutate(
        Method = Method |>
            str_replace_all("_", "-") |>
            str_replace("seurat-v3", "seurat_v3") |>
            str_replace("cell-ranger", "cell_ranger") |>
            factor(
                levels = methods_order,
                labels = method_names[methods_order]
            ),
        Dataset = factor(
            Dataset,
            levels = c("HLCA", "HLCAImmune", "HLCAEpithelial"),
            labels = c("Full", "Immune", "Epithelial")
        )
    )
 
markers_overlap_plot <- ggplot(marker_overlaps_plotting) +
    geom_tile(aes(x = Label, y = Method, fill = Prop)) +
    scale_fill_viridis_c(option = "plasma", limits = c(0, 1)) +
    scale_x_discrete(expand = c(0, 0)) +
    scale_y_discrete(expand = c(0, 0), drop = FALSE) +
    facet_grid(Dataset ~ Lineage, scales = "free_x", space = "free_x") +
    labs(
        title = "Overlap with marker gene sets by lineage",
        fill = "Proportion of marker genes selected"
    ) +
    theme_features_pub() +
    theme(
        legend.position = "bottom",
        axis.title = element_blank(),
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        panel.grid = element_blank(),
        panel.spacing.x = unit(0, "cm")
    )

markers_overlap_plot
```

# Output

Save output files

```{r output}
save_figure_files(
    lineages_figure,
    here::here("analysis", "output", "lineages"),
    height = 11
)

save_figure_files(
    markers_overlap_plot,
    here::here("analysis", "output", "markers-overlap"),
    height = 10
)
```

# Session info {.unnumbered}

```{r session-info}
sessioninfo::session_info()
```
